Dirac neutrino mass generation from Majorana messenger

### 2019-09-20 draft.tex
Updates from Overleaf

\begin{abstract}    
  The radiative type-I seesaw have been already implemented to explain the lightness of Majorana neutrinos with both Majorana and Dirac heavy fermions, and the lightness of Dirac    neutrinos with                   Dirac heavy fermions. In this work we present a minimal implementation of the radiative type-I seesaw with light Dirac neutrinos and heavy Majorana fermions. An inert doublet and a complex singlet scalar complete the dark sector which is protected by an Abelian fermiophobic gauge symmetry that also forbids three level mass contributions for the full set of light neutrinos. A fermion vector-like extension of the model is also proposed where the light right-handed neutrinos can thermalize in the primordial plasma and the extra gauge boson can be directly produced at colliders.
In particular, the  current upper bound on $\Delta N_{\text{eff}}$ reported by PLANCK points to large ratios $M_{Z'}/g'\gtrsim 40\ \text{TeV}$ which can be competetive with collider constraint for $g'$ sufficiently large in the ballpark of the SM values, while future cosmic microwave background experiments may probe all the no minimal models presented here. 
\end{abstract}

@@

In fact, another evidence that the SM is not a complete theory is the
missing matter content of the universe, which is known as dark matter
(DM).
The main proposals that explain DM as a particle are given in
Ref.~\cite{Bertone:2004pz}.
However, there have been only gravitational evidence for the existence
of dark matter so far.
Without evidence of DM as a particle, there is not a clear path to pin
out the DM properties nor the possible heavier companions of some
extended dark sector. 
Linking the dark sector to other specific phenomenology allows to
reduce the arbitrariness in the model building.
In our construction, the dark sector is related to the heavy sector
responsible of the lightness of the neutrinos and the same symmetry
which guarantees the lightness of the Dirac neutrinos is the
responsible of the stability of the lightest dark particle (LDP).
Therefore, the number of specific models is quite restricted.

% We will make use of the scotogenic model~\cite{Ma:2006km}. In
% scotogenic model the neutrino masses and the DM have common
% explanation. The neutrino masses are generated by one-loop effects,
% where the candidates for DM are mediators in the loop.
%Conclusion:
In this work we propose a radiative Dirac type-I seesaw with Majorana
fermions as mediators in the loop, where the full set of mediators in the
loop can include a DM candidate.

The rest of the paper is organized as follows. In the next Section we present the model and study the scalar mass spectrum after spontaneous symmetry breaking. In Sec.~\ref{sec:Neutrinos} we present the radiative mechanism to generate Dirac neutrino masses and establish  the lepton flavor violation constraints. The different DM scenarios are discussed in Sec.~\ref{sec:DM}. Finally, in Sec.~\ref{sec:CosmoConstraints} we show the cosmological restrictions ($N_{\text{eff}}$) in a non-minimum model for different extra abelian symmetries.

\section{The model}
\label{sec:Model}
%
\begin{table}[t!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline  
    Fields     & $\operatorname{SU}(2)_L$ & $\operatorname{U}(1)_Y $ & $\operatorname{U}(1)_{\mathcal{D}}$ \\ \hline
    % $H $  & 2  1/2  &  0 \\
    $\eta$  & $\boldsymbol{2}$ & $1$  & $1$ \\
    $S$ & $\boldsymbol{1}$ & $0$  & $2$ \\
    $\sigma$ & $\boldsymbol{1}$ & $0$ & $3$ \\
    \hline
    % $Q_{L_{i}}$  & $(2,-1/6)$ & $0$ \\
    % $\overline{u_{R_{i}}}$ & $(1,-2/3)$ & $0$ \\
    % $\overline{d_{R_{i}}}$ & $(1,1/3)$ & $0$ \\
    % \hline
    % $L_i$  & $(2,-1/2)$ & $0$ \\
    % $\overline{e_{R_i}}$ & $(1,1)$ & $0$ \\
    $\nu_{Ri}$ & $\boldsymbol{1}$ & $0$ & $-4$\\
    $\nu_{R3}$ & $\boldsymbol{1}$ & $0$ & $5$\\
    $\psi_{R\alpha}$  & $\boldsymbol{1}$ & 0 & $1$ \\\hline
  \end{tabular}
  \caption{The new scalars and fermions with their respective charges. All the SM fields are neutral under the dark $\operatorname{U}(1)_{\cal {D}}$ gauge symmetry. }
    \label{tab:partcont}
\end{table}
%
We extend the standard model with a spontaneously broken Abelian gauge
symmetry which guarantees the total lepton number ($L$)
conservation. Only the new particles, including the light right-handed
neutrinos, are charged under this new
$\operatorname{U}(1)_{\mathcal{D}}$ dark gauge
symmetry~\cite{Campos:2017dgc,Bertuzzo:2018itn,Bertuzzo:2018ftf} to obtain an anomaly free theory. We
choose the new particle set such that the following dimension six
operator is realized at one-loop level
\begin{align}
  \label{eq:ld6}
  \mathcal{O}_{6D}=\frac{1}{\Lambda^2} \overline{L} \tilde{H} \nu_R S^2\,,
\end{align}
where $S$ is the singlet scalar field which spontaneously breaks the $\operatorname{U}(1)_{\cal {D}}$ symmetry needed to forbid the Dirac and Majorana neutrino mass terms at tree level.

@@

with $V(\omega) = \mu^{2}_{\omega} \omega^{\dagger} \omega + \lambda_{\omega} (\omega^{\dagger} \omega)^{2}$. It is worth to emphasize that $\operatorname{U}(1)_{\mathcal{D}}$ automatically allows for all the terms in eqs.~\eqref{eq:LagY} and~\eqref{eq:VHSetasigma} but other realizations will be checked later on.

@@

On the other hand, we assume that the heavy Majorana fermions are already in the diagonal basis in such a way their masses are $M_{\psi_\alpha}=\kappa_{\alpha\alpha}v_S/\sqrt{2}$, with $M_{\psi_1}<M_{\psi_2}<M_{\psi_3}$. Finally, the mass of the new gauge boson is given by $M_{Z'}=q_Sg^{\prime} v_S$.
%\begin{pmatrix}
  %\psi_1 &
  %\psi_2 &
  %\psi_3
%\end{pmatrix}^{\operatorname{T}}$$

@@

%\subsection{Fermion dark matter}
Since $\psi_{1}$ is a singlet under the SM gauge group its thermal relic density is controlled by the Yukawa couplings and $\operatorname{U}(1)_{\mathcal{D}}$ gauge interactions. 
The scenario where $\psi_{1}$ self-annihilates dominantly through $h_{i\alpha}$-mediated interactions resembles the very well known scotogenic model \cite{Ma:2006km},  
where sizable $h_{i\alpha}$ are required to reproduce the correct DM abundance, which in turn leads to a mild tension with experimental upper bounds on the rates for rare charged lepton decays~\cite{Kubo:2006yx,Sierra:2008wj,Ibarra:2016dlb}. 
%The phenomenology of this scenario has been extensively studied  \cite{Kubo:2006yx,Sierra:2008wj,Suematsu:2009ww,Gelmini:2009xd,Hambye:2009pw,Aoki:2010tf,Schmidt:2012yg,Ma:2012if,Kashiwase:2012xd,Kashiwase:2013uy,Racker:2013lua,Toma:2013zsa,Molinaro:2014lfa,Vicente:2014wga,Merle:2015gea,Merle:2015ica,Merle:2016scw,Hagedorn:2016dze,Lindner:2016kqk,Hessler:2016kwm,Ibarra:2016dlb}.
On the other hand, when the $Z'$ portal \cite{Langacker:2008yv,Arcadi:2013qia} is the main gate to visible sector, $\psi_1$ largely annihilates into neutrinos in such a way the observed DM abundance can be reproduced without entering in conflict with the DM searches, which follows from the fact that the $Z'$ does not couples to quarks or charged leptons
(see~\cite{Arcadi:2013qia,Alves:2013tqa,Alves:2015pea,Alves:2016cqf,Blanco:2019hah} for phenomenological studies on $Z'$-mediated Majorana DM). 
% check Ibarra papers
%also through the annihilation
%$\psi_{1} \psi_{1} \to l _{\alpha}^+ \overline{l}_{\beta}^- $, via 
%Yukawa coupling $h^{a}_{i}$. These Yukawa couplings are restricted by
%CLFV and neutrino masses. %But, Yukawa couplings $y^{a}_{j}$ is free.
%In the case where the mass of the other fermions and scalars is nearly
%degenerated to $\psi_{1} $, the processes of coannihilation can be
%relevant for the relic density calculation~\cite{Klasen:2013jpa}.

@@

With the aim of thermalize the right-handed neutrinos with the
primordial plasma in the early universe, we consider an extension of
the previous setup by introducing a general anomaly free Abelian gauge
symmetry with generation-independent charge assignments for the
standard model fermions, $\operatorname{U}(1)_{X}$.
% 
In the appendix we show that the solutions to the anomaly cancellation conditions
allow us to write the $X$-charges of the SM fields in terms
of two parameters~\cite{Appelquist:2002mw,Campos:2017dgc}, that we choose as
\begin{align}
  X(r,l)=r R - l Y\,,
\end{align}
where $R$ is the generator of $\operatorname{U}(1)_R$: the gauge Abelian symmetry where only the right-handed SM fermions have non-vanishing $X$-charges, $Y$ is the hypercharge, $l$ is the $X$-charge of the lepton doublets, and $r$ parametrize the contribution to the linear and mixed gauge-gravitational anomalies of any extra set of chiral fermions, as given in Eq.~\eqref{eq:anolam} of the Appendix.
In this way, if only extra vector-like fermions are allowed beyond the SM ($r=0$) the solution must be proportional to hypercharge.
%
Without lost of generality, we can write the solutions in terms of just one parameter~\cite{Jenkins:1987ue,Oda:2015gna,Okada:2018tgy}, that we choose to be $l$  after fix $r=1$. Then, as shown in the Appendix, the full family of solutions for a fixed $l$ can be obtained after  rescaling all the $X$-charges by a factor $r$. In particular, the fermiophobic~\cite{Campos:2017dgc,Bertuzzo:2018itn,Bertuzzo:2018ftf} solution used in the previous sections, $\operatorname{U}(1)_{\mathcal{D}}$, corresponds to the rescaling $r=0$ of the $\operatorname{U}(1)_R$ solution: $\mathcal{D}=X(0,0)$.

The one parameter solution is shown in  column $\operatorname{U}(1)_X$
of Table~\ref{tab:partcont3}.
In order to analyse the phenomenology, we fix $l$ to recover some already studied
Abelian gauge groups $X=B-L,R,D,G,$ as defined in Ref.~\cite{Campos:2017dgc}\footnote{We change $\operatorname{U}(1)_B$ for the more suitable name of $\operatorname{U}(1)_R$~\cite{Jana:2019mez}.}. The last column corresponds to the reescaling with $r=0$ of $\operatorname{U}(1)_R$.

@@

    \Delta N_{\text{eff}} = N_{\text{eff}} - N^{\text{SM}}_{\text{eff}} = N_{\nu_R} \left( \frac{T_{\nu_{R}}}{T_{\nu_{L}}} \right)^{4} = N_{\nu_R} \left( \frac{g(T^{\nu_{L}}_{\text{dec}})}{g(T^{\nu_{R}}_{\text{dec}})} \right)^{4}

@@

where $N_{\nu_R}$ is the neutrino number with same charges and $g(T)$ is the number of relativistic degrees of freedom at a temperature $T$ in the SM~\cite{Aghanim:2018eyx}. The decoupling temperature of the SM neutrinos is $ T^{\nu_{L}}_{\text{dec}} \approx 2.3 $ MeV, when $g(T^{\nu_{L}}_{ \text{dec}}) = 43/4$ corresponding to the three $\nu_{L}$, $e^{\pm} $ and the photon~\cite{Kolb:1990vq,Enqvist:1991gx}. The rate of interaction of the right-handed neutrinos with the SM is mediated by the gauge boson $Z^{\prime} $ and is given by~\cite{SolagurenBeascoa:2012cz}

@@

    \Gamma_{\nu_R} (T) &= n_{\nu_R}(T) \langle \sigma(\overline{\nu_{R}} \nu_{R} \to \overline{f} f) \upsilon \rangle \,, \nonumber\\

@@

The cross section for the case of a heavy mediator ($T^{\nu_R}_{\text{dec}} \ll M_{Z^{\prime}}$), with $s \ll M_{Z^{\prime}}$ and neglecting the fermion masses in the final state is given by~\cite{Barger:2003zh}

@@

The results for $\Delta N_{\text{eff}}$ as a function of $M_{Z^{\prime}}/g^{\prime}$ are displayed in Fig.~\ref{fig:Neff}. It follows that for small rartios $M_{Z'}/g'$  the variation in the number of relativistic degrees of freedom of the new species is large ($\Delta N_{\text{eff}}\gtrsim0.3$ corresponds to decoupling temperatures $T^{\nu_R}_{\text{dec}}\lesssim2$ GeV) and vice versa.
% The bound that come from Planck~\cite{Aghanim:2018eyx} is stronger than the bound of LEP~\cite{Heeck:2014zfa}. and even LHC, see below
It follows that for the $\operatorname{U}(1)_{G}$ model the Planck+BAO upper bound demands that $M_{Z^{\prime}}/g^{\prime} \gtrsim 40$ TeV while for the $\operatorname{U}(1)_{R}$ model a more stringent lower bound applies $M_{Z^{\prime}}/g^{\prime} \gtrsim 60$ TeV. 
%in order to satisfy the conditions of $N_{\text{eff}}$ for all the symmetries studied here.
On the other hand, it is remarkable the fact the future CMB experiments~\cite{Benson:2014qhw,Abitbol:2019nhf} have the potential to entirely probe all the not-so minimal models~\cite{Abazajian:2019oqj}.  
%Prospect on $N_{\text{eff}}$~\cite{Abazajian:2019oqj}

>>>>

\begin{abstract}    
  The radiative type-I seesaw have been already implemented to explain the lightness of Majorana neutrinos with both Majorana and Dirac heavy fermions, and the lightness of Dirac    neutrinos with                   Dirac heavy fermions. In this work we present a minimal implementation of the radiative type-I seesaw with light Dirac neutrinos and heavy Majorana fermions. An inert doublet and a complex singlet scalar complete the dark sector which is protected by an Abelian fermiophobic gauge symmetry that also forbids tree level mass contributions for the full set of light neutrinos. A fermion vector-like extension of the model is also proposed where the light right-handed neutrinos can thermalize in the primordial plasma and the extra gauge boson can be directly produced at colliders.
In particular, the  current upper bound on $\Delta N_{\text{eff}}$ reported by PLANCK points to large ratios $M_{Z'}/g'\gtrsim 40\ \text{TeV}$ which can be competitive with collider constraint for $g'$ sufficiently large in the ballpark of the SM values, while future cosmic microwave background experiments may probe all the no minimal models presented here. 
\end{abstract}

@@

In fact, another evidence that the SM is not a complete theory is the
missing matter content of the Universe, which is known as dark matter
(DM).
The main proposals that explain DM as a particle are given in
Ref.~\cite{Bertone:2004pz}.
However, there has been only gravitational evidence for the existence
of dark matter so far.
Without evidence of DM as a particle, there is not a clear path to pin
out the DM properties nor the possible heavier companions of some
extended dark sector. 
Linking the dark sector to other specific phenomenology allows to
reduce the arbitrariness in the model building.
In our construction, the dark sector is related to the heavy sector
responsible of the lightness of the neutrinos and the same symmetry
that guarantees the lightness of the Dirac neutrinos is the
responsible of the stability of the lightest dark particle (LDP).
Therefore, the number of specific models is quite restricted.

% We will make use of the scotogenic model~\cite{Ma:2006km}. In
% scotogenic model the neutrino masses and the DM have common
% explanation. The neutrino masses are generated by one-loop effects,
% where the candidates for DM are mediators in the loop.
%Conclusion:

The rest of the paper is organized as follows. In the next Section we present the model and study the scalar mass spectrum after spontaneous symmetry breaking. In Sec.~\ref{sec:Neutrinos} we present the radiative mechanism that generates Dirac neutrino masses and establish  the lepton flavor violation constraints. The different resulting DM scenarios  are discussed in Sec.~\ref{sec:DM}, and in Sec.~\ref{sec:CosmoConstraints} we show the cosmological restrictions ($N_{\text{eff}}$) in a non-minimum model for different extra Abelian symmetries.

\section{The model}
\label{sec:Model}
%
\begin{table}[t!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline  
    Fields     & $\operatorname{SU}(2)_L$ & $\operatorname{U}(1)_Y $ & $\operatorname{U}(1)_{\mathcal{D}}$ \\ \hline
    % $H $  & 2  1/2  &  0 \\
    $\eta$  & $\boldsymbol{2}$ & $1$  & $1$ \\
    $S$ & $\boldsymbol{1}$ & $0$  & $2$ \\
    $\sigma$ & $\boldsymbol{1}$ & $0$ & $3$ \\
    \hline
    % $Q_{L_{i}}$  & $(2,-1/6)$ & $0$ \\
    % $\overline{u_{R_{i}}}$ & $(1,-2/3)$ & $0$ \\
    % $\overline{d_{R_{i}}}$ & $(1,1/3)$ & $0$ \\
    % \hline
    % $L_i$  & $(2,-1/2)$ & $0$ \\
    % $\overline{e_{R_i}}$ & $(1,1)$ & $0$ \\
    $\nu_{Ri}$ & $\boldsymbol{1}$ & $0$ & $-4$\\
    $\nu_{R3}$ & $\boldsymbol{1}$ & $0$ & $5$\\
    $\psi_{R\alpha}$  & $\boldsymbol{1}$ & 0 & $1$ \\\hline
  \end{tabular}
  \caption{The new scalars and fermions with their respective charges. All the SM fields are neutral under the dark $\operatorname{U}(1)_{\cal {D}}$ gauge symmetry. }
    \label{tab:partcont}
\end{table}
%
We extend the SM with a spontaneously broken Abelian gauge
symmetry which guarantees the total lepton number ($L$)
conservation. Only the new particles, including the  right-handed partners of the SM
neutrinos, are charged under this new
$\operatorname{U}(1)_{\mathcal{D}}$ dark gauge
symmetry~\cite{Campos:2017dgc,Bertuzzo:2018itn,Bertuzzo:2018ftf} to obtain an anomaly free theory. We
choose the new particle set such that the following dimension six
operator is realized at one-loop level
\begin{align}
  \label{eq:ld6}
  \mathcal{O}_{6D}=\frac{1}{\Lambda^2} \overline{L} \tilde{H} \nu_R S^2\,,
\end{align}
where $S$ is the singlet scalar field which spontaneously breaks the $\operatorname{U}(1)_{\cal {D}}$ symmetry needed to forbid the Dirac and Majorana neutrino mass terms at tree level.

@@

with $V(\omega) = \mu^{2}_{\omega} \omega^{\dagger} \omega + \lambda_{\omega} (\omega^{\dagger} \omega)^{2}$. It is worth to emphasize that $\operatorname{U}(1)_{\mathcal{D}}$ automatically allows for all the terms in Eqs.~\eqref{eq:LagY} and~\eqref{eq:VHSetasigma} but other realizations will be checked later on.

@@

On the other hand, we assume that the heavy Majorana fermions are already in the diagonal basis in such a way their masses are $M_{\psi_\alpha}=\kappa_{\alpha\alpha}v_S/\sqrt{2}$, with $M_{\psi_1}<M_{\psi_2}<M_{\psi_3}$. Finally, the mass of the new gauge boson is given by 
\begin{align}
    M_{Z'}=q_Sg^{\prime} v_S.
\end{align}
%\begin{pmatrix}
  %\psi_1 &
  %\psi_2 &
  %\psi_3
%\end{pmatrix}^{\operatorname{T}}$$

@@

%\subsection{Fermion dark matter}
Since $\psi_{1}$ is a singlet under the SM gauge group its thermal relic density is controlled by the Yukawa couplings and $\operatorname{U}(1)_{\mathcal{D}}$ gauge interactions. 
The scenario where $\psi_{1}$ self-annihilates dominantly through $h_{i\alpha}$-mediated interactions resembles the very well known scotogenic model \cite{Ma:2006km},  
where sizable $h_{i\alpha}$ are required to reproduce the correct DM abundance, which in turn leads to a mild tension with experimental upper bounds on the rates for rare charged lepton decays~\cite{Kubo:2006yx,Sierra:2008wj,Ibarra:2016dlb}. 
%The phenomenology of this scenario has been extensively studied  \cite{Kubo:2006yx,Sierra:2008wj,Suematsu:2009ww,Gelmini:2009xd,Hambye:2009pw,Aoki:2010tf,Schmidt:2012yg,Ma:2012if,Kashiwase:2012xd,Kashiwase:2013uy,Racker:2013lua,Toma:2013zsa,Molinaro:2014lfa,Vicente:2014wga,Merle:2015gea,Merle:2015ica,Merle:2016scw,Hagedorn:2016dze,Lindner:2016kqk,Hessler:2016kwm,Ibarra:2016dlb}.
On the other hand, when the $Z'$ portal \cite{Langacker:2008yv,Arcadi:2013qia} is the main gate to visible sector, $\psi_1$ largely annihilates into neutrinos in such a way the observed DM abundance can be reproduced without entering in conflict with the DM searches, which follows from the fact that the $Z'$ does not couple to quarks and charged leptons
(see~\cite{Arcadi:2013qia,Alves:2013tqa,Alves:2015pea,Alves:2016cqf,Blanco:2019hah} for phenomenological studies on $Z'$-mediated Majorana DM). 
% check Ibarra papers
%also through the annihilation
%$\psi_{1} \psi_{1} \to l _{\alpha}^+ \overline{l}_{\beta}^- $, via 
%Yukawa coupling $h^{a}_{i}$. These Yukawa couplings are restricted by
%CLFV and neutrino masses. %But, Yukawa couplings $y^{a}_{j}$ is free.
%In the case where the mass of the other fermions and scalars is nearly
%degenerated to $\psi_{1} $, the processes of coannihilation can be
%relevant for the relic density calculation~\cite{Klasen:2013jpa}.

@@

With the aim of thermalize the right-handed neutrinos with the
primordial plasma in the early Universe, we consider an modification of
the previous setup by considering a general anomaly free Abelian gauge
symmetry with generation-independent charge assignments for the
standard model fermions, $\operatorname{U}(1)_{X}$.
% 
In the Appendix we show that the solutions to the anomaly cancellation conditions
allow us to write the $X$-charges of the SM fields in terms
of two parameters~\cite{Appelquist:2002mw,Campos:2017dgc}, that we choose as
\begin{align}
  X(r,l)=r R - l Y\,,
\end{align}
where $R$ is the generator of $\operatorname{U}(1)_R$ (the gauge Abelian symmetry where only the right-handed SM fermions have non-vanishing $X$-charges), $Y$ is the hypercharge, $l$ is the $X$-charge of the lepton doublets, and $r$ parametrizes the contribution to the linear and mixed gauge-gravitational anomalies of any extra set of chiral fermions, as given in Eq.~\eqref{eq:anolam} of the Appendix.
In this way, if only extra vector-like fermions are allowed beyond the SM ($r=0$) the solution must be proportional to hypercharge.
%
Without lost of generality, we can write the solutions in terms of just one parameter~\cite{Jenkins:1987ue,Oda:2015gna,Okada:2018tgy}, that we choose to be $l$  after fix $r=1$. Then, as shown in the Appendix, the full family of solutions for a fixed $l$ can be obtained after  rescaling all the $X$-charges by a factor $r$. In particular, the fermiophobic~\cite{Campos:2017dgc,Bertuzzo:2018itn,Bertuzzo:2018ftf} solution used in the previous sections, $\operatorname{U}(1)_{\mathcal{D}}$, corresponds to the rescaling $r=0$ of the $\operatorname{U}(1)_R$ solution: $\mathcal{D}=X(0,0)$.

The one parameter solution is shown in  column $\operatorname{U}(1)_X$
of Table~\ref{tab:partcont3}.
In order to analyse the phenomenology, we fix $l$ to recover some already studied
Abelian gauge groups $X=B-L,R,D,G,$ as defined in Ref.~\cite{Campos:2017dgc}\footnote{We change $\operatorname{U}(1)_B$ for the more suitable name of $\operatorname{U}(1)_R$~\cite{Jana:2019mez}.}. The last column corresponds to the rescaling with $r=0$ of $\operatorname{U}(1)_R$.

@@

    \Delta N_{\text{eff}} = N_{\text{eff}} - N^{\text{SM}}_{\text{eff}} = N_{\nu_R} \left( \frac{T_{\nu_{R}}}{T_{\nu_{L}}} \right)^{4} = N_{\nu_R} \left( \frac{g(T^{\nu_{L}}_{\text{dec}})}{g(T^{\nu_{R}}_{\text{dec}})} \right)^{4},

@@

where $N_{\nu_R}$ is the number of right-handed neutrinos with the same $X$ charge and $g(T)$ is the number of relativistic degrees of freedom at a temperature $T$ in the SM~\cite{Aghanim:2018eyx}. The decoupling temperature of the SM neutrinos is $ T^{\nu_{L}}_{\text{dec}} \approx 2.3 $ MeV, when $g(T^{\nu_{L}}_{ \text{dec}}) = 43/4$ corresponding to the three $\nu_{L}$, $e^{\pm} $ and the photon~\cite{Kolb:1990vq,Enqvist:1991gx}. Since the interaction of the right-handed neutrinos with the SM is only mediated by the gauge boson $Z^{\prime} $, the corresponding rate can be cast as~\cite{SolagurenBeascoa:2012cz}

@@

    \Gamma_{\nu_R} (T) &= n_{\nu_R}(T) \langle \sigma(\overline{\nu_{R}} \nu_{R} \to \overline{f} f) \upsilon \rangle \, \nonumber\\

@@

The cross section for the case of a heavy mediator ($T^{\nu_R}_{\text{dec}} \ll M_{Z^{\prime}}$), with $s \ll M_{Z^{\prime}}$ and neglecting the fermion masses in the final state reads~\cite{Barger:2003zh}

@@

The results for $\Delta N_{\text{eff}}$ as a function of $M_{Z^{\prime}}/g^{\prime}$ are displayed in Fig.~\ref{fig:Neff}, where it can be observed that for small ratios $M_{Z'}/g'$  the variation in the number of relativistic degrees of freedom of the new species is large ($\Delta N_{\text{eff}}\gtrsim0.3$ corresponds to decoupling temperatures $T^{\nu_R}_{\text{dec}}\lesssim2$ GeV) and vice versa.
% The bound that come from Planck~\cite{Aghanim:2018eyx} is stronger than the bound of LEP~\cite{Heeck:2014zfa}. and even LHC, see below
It follows that for the $\operatorname{U}(1)_{G}$ model the Planck+BAO upper bound at $2\sigma$ (solid gray line) demands that $M_{Z^{\prime}}/g^{\prime} \gtrsim 40$ TeV while for the $\operatorname{U}(1)_{R}$ model a more stringent lower bound applies $M_{Z^{\prime}}/g^{\prime} \gtrsim 60$ TeV (these bounds become slightly weaker once the Planck+BAO+$H_0$ combination~\cite{Aghanim:2018eyx} is considered, but in such a case the BBN bound from the primordial abundances of light elements would rule~\cite{Pitrou:2018cgg}). 
%in order to satisfy the conditions of $N_{\text{eff}}$ for all the symmetries studied here.
In other words, bearing in mind that $M_{Z'}/g'=3v_S/2$ we have that the energy scale of the $\operatorname{U}(1)_{B-L}$ symmetry breaking must be at least $\sim26$ TeV.      
On the other hand, it is remarkable the fact the next generation of CMB experiments~\cite{Benson:2014qhw,Abitbol:2019nhf} has the potential to entirely probe all the not-so minimal models~\cite{Abazajian:2019oqj}.  
%Prospect on $N_{\text{eff}}$~\cite{Abazajian:2019oqj}
